graphsim: An R package for simulating gene expression data from graph structures of biological pathways

### 2020-06-10 paper/paper.Rmd
update intro and discussion

Gene networks provide insights into how cells are regulated, by mapping
regulatory interactions between target genes and transcription factors,
enhancers, and sites of epigenetic marks or chromatin structures
[@Barabasi2004; @Yamaguchi2007]. Inference of these regulatory
interactions for genomics investigations has the potential to radically
expand the range of candidate biological pathways to be further
explored, or to improve the accuracy of bioinformatics and functional
genomic analysis. A number of methods have already been developed to
utilise timecourse gene expression data [@Arner2015; @Yamaguchi2007]
using gene regulatory modules in state-space models and recursive vector
autoregressive models [@Hirose2008; @Shimamura2009]. Various approaches
to gene regulation and networks at the genome-wide scale have led to
novel biological insights [@Arner2015; @Komatsu2013]. However, inference
of regulatory networks has thus far relied on experimental validation or
resampling-based approaches to estimate the likelihood of specific
network modules being predicted [@Markowetz2007; @Hawe2019].

@@

Methodology and software {#sec:methods}
========================

@@

Biological pathways are of fundamental importance to understanding
molecular biology. In order to translate findings from genomics studies
into real-world applications such as improved healthcare, the roles of
genes must be studied in the context of molecular pathways. Here we
present a statistical framework to simulate gene expression from
biological pathways, and provide the \texttt{graphsim} package in R to generate these
simulated datasets. This approach is versatile and can be fine-tuned for
modelling existing biological pathways or for testing whether
constructed pathways can be detected by other means. In particular,
methods to infer biological pathways and gene regulatory networks from
gene expression data can be tested on simulated datasets using this
framework. The package also enables simulation of complex gene
expression datasets to test how these pathways impact on statistical
analysis of gene expression data using existing methods or novel
statistical methods being developed for gene expression data analysis.

>>>>

Gene networks provide insights into how cells are regulated, by mapping
regulatory interactions between target genes and transcription factors,
enhancers, and sites of epigenetic marks or chromatin structures
[@Barabasi2004; @Yamaguchi2007]. Inference of these regulatory
interactions for genomics investigations has the potential to radically
expand the range of candidate biological pathways to be further
explored, or to improve the accuracy of bioinformatics and functional
genomic analysis. A number of methods have already been developed to
utilise timecourse gene expression data [@Arner2015; @Yamaguchi2007]
using gene regulatory modules in state-space models and recursive vector
autoregressive models [@Hirose2008; @Shimamura2009]. Various approaches
to gene regulation and networks at the genome-wide scale have led to
novel biological insights [@Arner2015; @Komatsu2013]. However, inference
of regulatory networks has thus far primarily relied on experimental
validation or resampling-based approaches to estimate the likelihood
of specific network modules being predicted [@Markowetz2007; @Hawe2019].

Simulated datasets with pathway structure are of particular interest for
benchmarking regulatory network inference techniques and methods
being developed for genomics data with complex biological 
interactions [@Schaffter2011, @Saelens2019]. Dynamical models using differential equations have been employed,
such as by GeneNetWeaver [@Schaffter2011], to generate simulated datasets
specifically for benchmarking gene regulatory network inference techniques.
There is renewed interest in modelling biological pathways and simulating
data for benchmarking due to the emergence of single-cell genomics
technologies and the plethora of bioinformatics techniques developed
to use this data [@Zappia2017, @Saelens2019].

Splatter [@Zappia2017] was developed to model single-cell data and 
SERGIO [@Dibaeinia2019] and dyngen [@Cannoodt2020] build on this to
include gene regulatory networks and multimodality respectively.

These methods have been designed based on known or modelled reaction
states and added stochasticity to determinisitic relationships.
However, it is computationally-intensive to model these reactions
at scale over iterations in benchmarking. In some cases, it is only
necessary to model the statistical variability of RNA-Seq data,

GeneNetWeaver @Schaffter2011 - benchmarking of gene regulatory network inference methods
Splatter @Zappia2017 - simulated single-cell RNA-Seq data (gamma-poisson)

dyngen @Cannoodt2020 - single-cell multimodal -  Gillespie’s stochastic simulation at single-molecule level

SERGIO @Dibaeinia2019 - single-cell gene expression (noise with poisson)

@@

Methodology and software {#sec:methods}
========================d

@@

Biological pathways are of fundamental importance to understanding
molecular biology. In order to translate findings from genomics studies
into real-world applications such as improved healthcare, the roles of
genes must be studied in the context of molecular pathways. Here we
present a statistical framework to simulate gene expression from
biological pathways, and provide the \texttt{graphsim} package in R 
to generate these simulated datasets. This approach is versatile and
can be fine-tuned for modelling existing biological pathways or for
testing whether constructed pathways can be detected by other means.
In particular, methods to infer biological pathways and gene regulatory
networks from gene expression data can be tested on simulated datasets
using this framework. The package also enables simulation of complex gene
expression datasets to test how these pathways impact on statistical
analysis of gene expression data using existing methods or novel
statistical methods being developed for gene expression data analysis.
This approach is intended to be applied to bulk gene expression data
but could in principle be adapted to modelling single-cell or
different modalities such as epigenomics data.
